---
title: "CanEpiRisk"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CanEpiRisk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CanEpiRisk)
```


## Overview

Quantifying cancer risks associated with ionizing radiation exposure is a central concern in radiation epidemiology and public health, particularly when estimating lifetime risks across populations and exposure scenarios. `CanEpiRisk` is a comprehensive R package designed to facilitate radiation-associated lifetime risk calculations. The package provides an integrated framework for importing exposure and demographic data, transferring specified risk models—such as **excess relative risk (ERR)** and **excess absolute risk (EAR)** and computing key risk measures, including **cumulative excess risk (CER)** and **years of life lost (YLL)**. By offering a unified interface for evaluating risks under various exposure scenarios and across diverse population settings, **CanEpiRisk** streamlines the workflow for radiation risk assessment, enhances analytical reproducibility, and enables transparent comparison of model-based estimates.


## Quick Start

[Using/Specifying reference data](https://KyojiFurukawa.github.io/CanEpiRisk/articles/CanEpiRisk_ReferenceData.html).

[Using/Specifying risk models](https://KyojiFurukawa.github.io/CanEpiRisk/articles/CanEpiRisk_Models.html)

[Calculating risks](https://KyojiFurukawa.github.io/CanEpiRisk/articles/CanEpiRisk_RiskCalculation.html).



## Quick examples

```   
# Cumulative excess risk (CER) calucations 
# (1) All solid cancer mortality, Region-1, female, 0.1Gy at age 15, followed up to age 100, LSS linear ERR
exp1 <- list( agex=5, doseGy=0.1, sex=2 )   # exposure scenario
ref1 <- list( baseline=Mortality[[1]]$allsolid,        # baseline rates
             mortality=Mortality[[1]]$allcause )       # all-cause mortality
mod1 <- LSS_mortality$allsolid$L                       # risk model
opt1 <- list( maxage=100, err_wgt=1, n_mcsamp=10000 )  # option
CER(  exposure=exp1, reference=ref1, riskmodel=mod1, option=opt1 ) * 10000 # cases per 10,000

# (2) Leukaemia incidence, Region-4, male, 6.7(100/15)mGy at ages 30-45, followed up to age 60, LSS LQ EAR
exp2 <- list( agex=30:44+0.5, doseGy=rep(0.1/15,15), sex=1 )
ref2 <- list( baseline=Incidence[[4]]$leukaemia,       # baseline rates
             mortality=Mortality[[4]]$allcause )       # all-cause mortality rates
mod2 <- LSS_incidence$leukaemia$LQ                     # risk model
opt2 <- list( maxage=60, err_wgt=0, n_mcsamp=10000)    # option
CER(  exposure=exp2, reference=ref2, riskmodel=mod2, option=opt2 ) * 10000 # cases per 10,000
```

