---
title: "Using/Specifying Risk Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CanEpiRisk: Using or Specifying Risk Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installation

```{r setup}
library(CanEpiRisk)
```

## Overview

A risk model used for risk calculations with `CanEpiRisk` package may be a predefined model from either LSS mortality risk model object (*LSS_mortality*) or LSS cancer incidence model object (`LSS_mortality`). Otherwise, the user can specify any risk model by a list object with a certain format.

## (1) Predefined Risk Models

ERR and EAR risk models for cancer mortality and incidence derived from recent Life Span Study of Japanese atomic-bomb studies are available in predefined objects; `LSS_mortality` and `LSS_incidence`. Each risk model for an endpoint (i.e., a site-specific cancer incidence or mortality) is a list object which contains a vector of parameter estimates `para`, a matrix of variance-covariance matrix of parameter estimates `var` and a function to calculate the risk `f`.

```{r}
# Mortality
names(LSS_mortality)   # Sites for which LSS mortality risk models are available
names(LSS_mortality$allsolid)    # Available dose response models
LSS_mortality$allsolid$L$err     # Linear ERR model for all solid cancer motality
LSS_mortality$allsolid$L$ear     # Linear EAR model for all solid cancer motality

LSS_mortality$leukaemia$LQ$err   # Linear-quadratic ERR model for leukaemia motality
LSS_mortality$lung$L$err         # Linear EAR model for lung cancer motality

# Plotting LSS all solid cancer mortality risk model
plot_riskmodel( rm=LSS_mortality$allsolid$L, title="LSS all solid cancer mortality, Linear",  leg_pos=c(0.4, 0.95) )

# Incidence
names(LSS_incidence)   # Sites for which LSS incidence risk models are available
names(LSS_incidence$allsolid)    # Available dose response models
LSS_incidence$allsolid$L$err     # Linear ERR model for all solid cancer incidence
LSS_incidence$allsolid$L$ear     # Linear EAR model for all solid cancer incidence
# Plotting LSS all solid cancer incidence risk model
plot_riskmodel( rm=LSS_incidence$allsolid$L, title="LSS all solid cancer incidence, Linear",  leg_pos=c(0.4, 0.95) )
  
```

## (2) Specifying Risk Models

A risk model may be specified by a list object which has the same structure as in an LSS risk model object (`LSS_mortality` or `LSS_incidence`). As an example, the risk models derived from the INWORKS cohort for mortality from all solid cancer (Richardson et al., 2015) and from leukaemia (Leuraud et al., 2015) can be specified as follows. Note that the uncertainty in the only risk model parameter (ERR/Gy) is specified by the 95% confidence interval, instead of the variance-covariance matrix in the LSS risk models objects.

```{r}
INWORKS_mortality <- NULL
INWORKS_mortality$allsolid <- NULL

INWORKS_mortality$allsolid$L <- list(
     err=list(
       para=c(0.47),                  # ERR/Gy=0.47 (90% CI: 0.18,0.79)  
       ci= c(0.1392403, 0.8521128),   #  95% CI coverted from 90%CI by Weibull approx.
       f=function (beta, data, lag=10) {
           beta[1] * data$dose  * (data$age - data$agex >= lag )
       }
       ),
      ear=list(    # dummry object
       para=c(4.8/10000),
       ci= c(0.1068428, 12.5703871)/10000,
       f=function (beta, data, lag=10) {
           beta[1] * data$dose  * (data$age - data$agex >= lag )
       } 
       )
 
  )

INWORKS_mortality$leukaemia$L <- list(
     err=list(
       para=c(2.96),                         # ERR/Gy=2.96 (90% CI: 1.17, 5.21) 
       ci= c(0.8664, 5.6940),
       f=function (beta, data, lag=2) {
           beta[1] * data$dose  * (data$age - data$agex >= lag )
       }
       ),
      ear=list(    # dummry object
       para=c(2.25/10000),
       ci=c(0.5064054, 4.4914628)/10000,
       f=function (beta, data, lag=2) {
           beta[1] * data$dose  * (data$age - data$agex >= lag )
       } 
       )
  )

INWORKS_mortality


# Plotting solid cancer mortality risk, male, 
#          6.7(100/15)mGy at ages 30-45, followed up to age 60, LSS-L, LSS-LQ and INWORKS models

exp4 <- list( agex=30:44+0.5, doseGy=rep(0.1/15,15), sex=1 )   # Exposure scenario
opt4_err <- list( maxage=90, err_wgt=1 )    # for ERR transfer
opt4_ear <- list( maxage=90, err_wgt=0 )    # for EAR transfer

lss_err_L  <- Comp_Exrisk( exposure=exp4, riskmodel=LSS_mortality$allsolid$L,     option=opt4_err )
lss_err_LQ <- Comp_Exrisk( exposure=exp4, riskmodel=LSS_mortality$allsolid$LQ,    option=opt4_err )
inw_err_L  <- Comp_Exrisk( exposure=exp4, riskmodel=INWORKS_mortality$allsolid$L, option=opt4_err )

lss_ear_L  <- Comp_Exrisk( exposure=exp4, riskmodel=LSS_mortality$allsolid$L,     option=opt4_ear, per=10^4 )
lss_ear_LQ <- Comp_Exrisk( exposure=exp4, riskmodel=LSS_mortality$allsolid$LQ,    option=opt4_ear, per=10^4 )
inw_ear_L  <- Comp_Exrisk( exposure=exp4, riskmodel=INWORKS_mortality$allsolid$L, option=opt4_ear, per=10^4 )

plot( c(30,90), c(0,0.1), type="n", ylab="Excess relative rate", xlab="age (years)" )
lines( lss_err_L, lty=2 )
lines( lss_err_LQ, lty=3 )
lines( inw_err_L )

plot( c(30,90), c(0,5), type="n", ylab="Excess absolute rate (cases per 10,000 person years)", xlab="age (years)" )
lines( lss_ear_L, lty=2 )
lines( lss_ear_LQ, lty=3 )
lines( inw_ear_L )


```

## References

Richardson, D.B., E. Cardis, R.D. Daniels et al. Risk of cancer from occupational exposure to ionising radiation: retrospective cohort study of workers in France, the United Kingdom, and the United States (INWORKS). BMJ 351: h5359 (2015).

Richardson, D.B., K. Leuraud, D. Laurier et al. Cancer mortality after low dose exposure to ionising radiation in workers in France, the United Kingdom, and the United States (INWORKS): cohort study. BMJ 382: e074520 (2023).

Leuraud, K., D.B. Richardson, E. Cardis et al. Ionising radiation and risk of death from leukaemia and lymphoma in radiation-monitored workers (INWORKS): an international cohort study. Lancet Haematol 2(7): e276-281 (2015).
