---
title: "Occupational Scenario: Radiation Risk Calculation Using CanEpiRisk"
author: "Furukawa"
output:
  html_document:
    toc: true
    number_sections: true
    theme: readable
---

# Introduction

This document reproduces **Scenario 4 (Occupational Scenario)** from the UNSCEAR CanEpi Report using the **CanEpiRisk** R package.

Scenario assumptions:

- Population: Male worker  
- Exposure period: Ages 20–45  
- Total dose: 100 mGy (0.1 Gy)  
- Outcome: Solid cancer mortality  
- Risk models: LSS and INWORKS  
- Risk metric: CER  
- Risk accumulated to: Age 60  

---

# Install and load CanEpiRisk

```{r}
# Install CanEpiRisk from GitHub if not installed
if (!require("CanEpiRisk")) {
  install.packages("devtools")
  devtools::install_github("KyojiFurukawa/CanEpiRisk")
}

library(CanEpiRisk)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
```

---

# Define Scenario 4 exposure

```{r}
exp4 <- list(
  agex   = 30:44 + 0.5,
  doseGy = rep(0.1/15, 15),
  sex    = 1
)
exp4
```

---

# Exposure profile plot

```{r}
df_exp <- data.frame(
  age  = exp4$agex,
  dose = exp4$doseGy
)

ggplot(df_exp, aes(x = age, y = dose)) +
  geom_col(fill = "steelblue") +
  labs(title = "Exposure Profile (Scenario 4)",
       x = "Age at Exposure",
       y = "Dose per Year (Gy)") +
  theme_minimal()
```

---

# Reference data (Region 1)

```{r}
ref4_1 <- list(
  baseline = Mortality[[1]]$allsolid,
  mortality = Mortality[[1]]$allcause
)
ref4_1
```

---

# Risk models

```{r}
rm_LSS_L  <- LSS_mortality$allsolid$L
rm_LSS_LQ <- LSS_mortality$allsolid$LQ
rm_INW_L  <- INWORKS_mortality$allsolid$L

rm_LSS_L
```

---

# Calculation options

```{r}
opt4_err <- list(maxage = 60, err_wgt = 1, n_mcsamp = 10000)
opt4_ear <- list(maxage = 60, err_wgt = 0, n_mcsamp = 10000)

opt4_err
```

---

# CER calculations (per 10,000)

```{r}
cer_LSS_L_err  <- CER(exp4, ref4_1, rm_LSS_L,  opt4_err) * 10000
cer_LSS_L_ear  <- CER(exp4, ref4_1, rm_LSS_L,  opt4_ear) * 10000
cer_LSS_LQ_err <- CER(exp4, ref4_1, rm_LSS_LQ, opt4_err) * 10000
cer_LSS_LQ_ear <- CER(exp4, ref4_1, rm_LSS_LQ, opt4_ear) * 10000
cer_INW_err    <- CER(exp4, ref4_1, rm_INW_L,  opt4_err) * 10000
cer_INW_ear    <- CER(exp4, ref4_1, rm_INW_L,  opt4_ear) * 10000

cer_LSS_L_err
```

---

# CER summary table

```{r}
cer_results <- data.frame(
  Model = c("LSS Linear ERR", "LSS Linear EAR",
            "LSS LQ ERR", "LSS LQ EAR",
            "INWORKS ERR", "INWORKS EAR"),
  CER_mean = c(cer_LSS_L_err$mean,
               cer_LSS_L_ear$mean,
               cer_LSS_LQ_err$mean,
               cer_LSS_LQ_ear$mean,
               cer_INW_err$mean,
               cer_INW_ear$mean)
)

kable(cer_results, caption = "CER Results for Scenario 4 (per 10,000)")
```

---

# CER curves as a function of age

```{r}
ages <- 30:60

calc_curve <- function(model, opt) {
  sapply(ages, function(a) {
    exp_tmp <- exp4
    exp_tmp$agex <- a
    CER(exp_tmp, ref4_1, model, opt)$mean * 10000
  })
}

df_curve <- data.frame(
  age        = ages,
  LSS_L_ERR  = calc_curve(rm_LSS_L,  opt4_err),
  LSS_L_EAR  = calc_curve(rm_LSS_L,  opt4_ear),
  LSS_LQ_ERR = calc_curve(rm_LSS_LQ, opt4_err),
  LSS_LQ_EAR = calc_curve(rm_LSS_LQ, opt4_ear),
  INW_ERR    = calc_curve(rm_INW_L,  opt4_err),
  INW_EAR    = calc_curve(rm_INW_L,  opt4_ear)
)

df_long <- pivot_longer(df_curve, -age,
                        names_to = "Model",
                        values_to = "CER")

ggplot(df_long, aes(x = age, y = CER, color = Model)) +
  geom_line(size = 1.1) +
  labs(title = "CER Curves for Scenario 4",
       x = "Age",
       y = "CER per 10,000") +
  theme_minimal()
```

---

# Interpretation

- LSS Linear ERR model: ~6–7 per 10,000  
- INWORKS ERR model: ~10 per 10,000  
- LSS LQ models: lower CER due to curvature  
- EAR models: generally higher CER  

These values represent the **excess lifetime risk attributable to 100 mGy occupational exposure**, accumulated to age 60.

---

# Summary

- CanEpiRisk provides unified ERR/EAR risk calculations  
- Scenario 4 represents realistic chronic occupational exposure  
- CER tables and curves help compare risk across models  

